Sequential rank agreement methods for comparison of ranked lists

### 2017-05-23 latex/superranker-manuscript.tex
Minor tweaks

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete or
  censored. The approach can identify changepoints in the agreement
  of the lists, and the sequential changes of agreement as a function
  of the depth of the lists can be compared graphically to a
  permutation based reference set or evaluated using analytical asymptotic
  results.  The usefulness of these tools are illustrated using gene
  rankings, and using data from two Danish ovarian cancer studies
  where we assess the within and between agreement of different
  statistical classification methods.
\end{abstract}

@@

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the rankings is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all depths $d$.  The sequential rank
agreement is equivalent to the pooled variance of the items found in
$S(d)$, and it can be interpreted as the average distance of the
individual rankings of the lists from the average ranking for each of
the items we have seen until depth $d$.  For an easier interpretation
we suggest to plot $\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding
to the pooled standard deviation.

@@

The sequential rank agreement curve plotted in Figure~\ref{fig:example1}
roughly shows how many ranks apart we on average expect the genes found in the
top $d$ part of the lists to be for each depth $d$.  It can be seen
that the sequential rank agreement limits are better towards the top
of the lists (smaller values on the $y$ axis corresponds to
\emph{better} agreement) than towards the tails of the lists. It is clear
from the curve in Figure~\ref{fig:example1} that there is a
substantial drop in agreement even after the first depth. Thus, if we
were to restrict attention to a small set of predictors then we would
focus on predictors 2124, 829, and 378 as seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{Top 10 list of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & Welsh's $t$ & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data (black
   line). Right
   panels: Corresponding sequential rank agreement for the same data
   but where only the top 20 ranked items are available and the rest
   of the items have been censored in each list. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

@@

\subsection{Analysis of partial/censored lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering. Incomplete lists also occur if the analyst decides
to censor all lists at depth $d$ (\ie, top $d$ lists) such that the
measurement of agreement is restricted to the first $d$ elements of
each list, or when the analyst only wants to consider the ranking of
items that have been found to be statistically significant.

Sequential rank agreement can be generalized to partial lists in the
following way. Let $\Lambda_l\subset X$ be the subset of $d_l$ items
that have been ranked highest in list $l$.  The case where all lists are
censored at the same depth $d$ corresponds to $d_1=\cdots=d_L=d$. For
incomplete/censored lists the rank function becomes
\begin{equation}
	\widetilde R_l(X_p) = \begin{cases}
	     \{R_l(X_p)\} & \text{for } X_p\in \Lambda_l\\
	     \{d_l+1,\dots,P\} & \text{for} X_p \not\in \Lambda_l
        \end{cases}
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of censored lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(\eg, the mean of the unassigned ranks) to the censored items since
that would result in \emph{less} variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are censored in multiple
lists.

@@

\begin{algorithm}
\caption{Sequential rank agreement algorithm for censored lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Censored rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each censored list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=X\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the most
interesting items are found in the top of the lists, and 2) that the
censored rankings provide so little information that it is reasonable
to assume that they can be represented by a random order. The first
assumption is justifiable because we have already accepted that it is
reasonable to rank the items in the first place. The second assumption
is fair in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of censored items is
low. Generally, without additional prior knowledge about the
distribution of the censored items it seems reasonable to restrict the
attention of the sequential rank agreement to depths smaller than
$\max(d_1, \ldots, d_L)$ where at least one list provides an
uncensored item.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For partially ranked lists we also
expect a changepoint around the depth where the lists are
censored. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

@@

The first reference hypothesis is
\begin{eqnarray}
H_0  & : &  \text{The list rankings correspond to completely randomly}\label{eq:permutationHypothesis}\\
       &  & \text{permuted lists}\nn
\end{eqnarray}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods are completely
independent.

Alternatively, we can remove the restriction on the independence among
the methods by only requiring that there is no information in the
ranking but that the rankings are all based on applying the
method/approaches to the same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
censored/partially observed case we first permute the items
$X_1,\dots,X_P$ and then censor list $l$ at $d_l$ (\ie, each list so
it has the same number of observed rankings as was observed for list
$l$ in the original dataset). The sequential rank agreement curve from
the the original lists can then be compared to, say, the pointwise
95\% quantiles of the observed rank agreements obtained under $H_0$.

@@

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. The censored
data also suggests that there may be at most 1 or 2 ranked items
towards the top of the lists that yield a result better than what
would be expected (the bottom-right plot).

@@

Corollary \ref{thecor} indicates that we can use the a threshold
function $\widehat{q}_L$ estimated under the null hypothesis as
discussed in the previous section as a limiting threshold function for
inferring the depth $d$, where the observed sequential rank agreement
first crosses the threshold of the null threshold, \ie, the depth
until which the observed ranked lists are in better agreement than
expected under the null hypothesis. In that sense the threshold
function serves the same role as the limits of agreement in method
comparison studies, except that the threshold funciton is not constant
but can accommodate the changing nature of the number of items used
for the computation of the equential rank agreement for a given depth.

@@

The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce censored lists depending on
the value of the penalty parameter.

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we see that a large
proportion of them are numerically very close to zero and have almost
equal absolute value. This is a general feature of the Ridge models in
this dataset and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient values
below a given threshold thereby introducing an artificial censoring of the lists.
For the Ridge models tuned with the deviance criterion, Figure \ref{fig:app2} (left) shows the
the sequential rank agreement where for each of the 1000 trained
models the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated
using Algorithm \ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (right panel) is shown for comparison.
Even though the number of censored predictors is very small compared to the total
number of predictors, the effect on the sequential rank agreement is
substantial and with the artificial censoring the shape of the curves
is as expected, starting low and then increasing.

Looking at the agreement curves for the Lasso models in Figure \ref{fig:app1}
(right) we clearly see the effect of the sparsity inducing penalization giving
rise to censored lists. These curves were similarly calculated using \ref{sra-algorithm}
and $1000$ random permutations. Under the deviance optimization criterion the
median number of non-zero coefficients was 33 (range 16 to 50) and for the
class accuracy criterion 14 (range 4 to 56). These values correspond to
the list depths where the agreement curves become constant as a result of
the subsequent censoring.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.49\textwidth]{pics/ridgeCensoring}%
\includegraphics[width=.49\textwidth]{pics/aucPlot}
\end{center}
\caption{Left panel: Sequential rank agreement for Ridge regression
  obtained by artificially censoring predictors when their absolute
  coefficient values are lower than the 0.1\% quantile. Right panel:
  Box plots of AUC values across the $1000$ sub-samples with respect
  to the known class labels of the DACOVA data.}
 \label{fig:app2}
\end{figure}

@@

Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (except for the AUC criterion), and poorer agreement for
ranking predictors. This reason for the latter is the high
auto-correlation between the intensities in the mass spectra which
leads to collinearity issues in the regression models. It is
well-known that variable selection with the Lasso penalty does not
preform very well when the predictors are highly correlated. The
collinearity does, however, not affect the agreement of the risk
predictions, as seen on the left panel of Figure \ref{fig:app1}, since
it is not so important which specific variable that gets selected from
a group of highly correlated predictors when the purpose is risk
predictions.

@@

We present results of a simulation study where we investigated the
small sample properties of the sequential rank agreement and compared
it with the topK method \citep{hall:schi:2012} with respect to list
depth agreement. For the sequential rank agreement we use the function
\texttt{sra} of the \texttt{R} package \texttt{SuperRanker} with the
\texttt{"mad"} option (median absolute distance). To define the depth
of agreement we set a constant threshold function to an integer \(q\)
and report the first crossing point, i.e., the smallest list depth
where sra exceeds \(q\). For topK we use the function
\texttt{j0.multi} which is implemented in the \texttt{R} package
\texttt{TopKLists}. Specifically, we set the tuning parameter
\texttt{v} of \texttt{j0.multi} to the value 6 and the window
parameter \texttt{d} to \(q\) and report the output parameter
\texttt{maxK} as the depth of agreement. Thus, to make this comparison
we assume that the first crossing point of sra and the result of the
topK method measure the same underlying feature.

@@

\section{Discussion}
In this article we address the problem of comparing ranked lists of
the same items. Our proposed method can handle both the situation
where the underlying data to generate the ranked lists are available
and the situation where the only available data is the actual ranked
lists. In addition, censored ranked lists where only the ranks of the
top $k$ ranked items are known can be accommodated as well. The
proposed agreement measure can be interpreted as the average distance
between an item's rank and the average rank assigned to that item
across lists. Thus the measure determines how well the lists agree on
the ranks of a specific item.

@@

Finally, we have --- whenever possible --- used all available ranks
from the lists.  We could choose to censor the rankings of all the
items that were deemed not to be significant if information on both
rankings and the corresponding evidence for significance was
available. That would ensure that there would be put less emphasis on
the agreement of the non-significant items and it would be easier to
identify a change in agreement among the items that were deemed to
be relevant. In our application section we have successfully introduced
such an artificial censoring for the predictor rankings obtained with
ridge regression.

>>>>

\begin{abstract}
  The comparison of alternative rankings of a set of items is a
  general and common task in applied statistics. Predictor variables
  are ranked according to magnitude of association with an outcome,
  prediction models rank subjects according to the personalized risk
  of an event, and genetic studies rank genes according to their
  difference in gene expression levels. This article constructs
  measures of agreement among two or more ordered lists. We use the
  standard deviation of the ranks to define a measure of agreement
  that both provides an intuitive interpretation and can be applied to
  any number of lists even if some or all are incomplete. The approach
  can identify changepoints in the agreement of the lists, and the
  sequential changes of agreement as a function of the depth of the
  lists can be compared graphically to a permutation based reference
  set or evaluated using analytical asymptotic results.  The
  usefulness of these tools are illustrated using gene rankings, and
  using data from two Danish ovarian cancer studies where we assess
  the within and between agreement of different statistical
  classification methods.
\end{abstract}

@@

Values of \textrm{sra} close to zero suggests that the lists agree on
the rankings while larger values suggest disagreement.  Just like in
method comparison studies we prefer smaller limits of agreement
because that means that the differences between the rankings is
small. If the ranked lists are identical then the value of sequential
rank agreement will be zero for all position or depths $d$ down the
lists.  The sequential rank agreement is equivalent to the pooled
variance of the items found in $S(d)$, and it can be interpreted as
the average distance of the individual rankings of the lists from the
average ranking for each of the items we have seen until depth $d$.
For an easier interpretation we suggest to plot
$\sqrt{\widehat{\textrm{sra}}_L(d)}$ corresponding to the pooled
standard deviation.

@@

The sequential rank agreement curve shown in the two left plots of
Figure~\ref{fig:example1} roughly indicates the average distance in
ranks for the genes found among the first $d$ positions in the $4$
lists.  It can be seen that the sequential rank agreement limits are
better towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. It is clear from the curve in Figure~\ref{fig:example1} that
there is a substantial drop in agreement (higher sra) even after the
first depth. Thus, if we were to restrict attention to a small set of
predictors then we would focus on predictors 2124, 829, and 378 as
seen in Table~\ref{tab1}.

% latex table generated in R 3.2.3 by xtable 1.8-2 package
% Fri Apr  8 10:22:58 2016
\begin{table}[tb]
\centering
\caption{List of ranked results from the Golub data. Numbers indicate the predictor/gene for the given ranking and method. Only the top 10 ranks are shown in the table.} 
\label{tab1}
\begin{tabular}{rrrrr}
  \hline
Ranking & Welsh's $t$ & LogReg & ElasticNet & MIC \\ 
  \hline
1 & 2124 & 2124 & 829 & 378 \\ 
  2 & 896 & 896 & 2124 & 829 \\ 
  3 & 2600 & 829 & 2198 & 896 \\ 
  4 & 766 & 394 & 1665 & 1037 \\ 
  5 & 829 & 766 & 1920 & 2124 \\ 
  6 & 2851 & 2670 & 1042 & 808 \\ 
  7 & 703 & 2939 & 808 & 108 \\ 
  8 & 2386 & 2386 & 849 & 515 \\ 
  9 & 2645 & 1834 & 937 & 2670 \\ 
  10 & 2002 & 378 & 1995 & 2600 \\ 
   \hline
\end{tabular}
\end{table}

@@

\begin{figure}[htbp]
   \begin{center}
 \includegraphics[width=.49\textwidth]{paper-fig1a}
 \includegraphics[width=.49\textwidth]{paper-fig2a}
 \includegraphics[width=.49\textwidth]{paper-fig1b}
 \includegraphics[width=.49\textwidth]{paper-fig2b}
 \end{center}
 \caption{Left panels: Sequential rank agreement for 4 different
   analysis methods applied to the 3051 genes in the Golub data (black
   line). Right panels: Corresponding sequential rank agreement for
   the same data but where only the top 20 ranked items are available
   and the rank of the remaining items are not available. The blue and
   red areas correspond to the independent and randomized reference
   hypothesis areas, respectively. The bottom plots are identical to
   the corresponding top plots but have been zoomed in on the first
   part of the $x$ axis.}
 \label{fig:example1}
 \end{figure}

@@

\subsection{Analysis of incomplete lists}
Incomplete or partial lists are a common occurrence that arise, for
example, in case of missing data (items), when comparing top $d$ list
results from publications, or when some methods only rank a subset of
the items. For example, penalized regression based on the Lasso
provides a sparse set of predictors that have non-zero
coefficients. There is no obvious ordering of the set of predictors
whose coefficient has been shrunk to zero and thus we end up with a
partial ordering.  Incomplete lists also occur if the analyst
restricts attention to the ranks of items that have been found to be
statistically significant.

Sequential rank agreement can be generalized to incomplete lists in
the following way. Let $\Lambda_l\subset X$ be the subset of $d_l$
items that have been ranked highest in list $l$.  The case where all
lists are incomplete at the same depth $d$ corresponds to
$d_1=\cdots=d_L=d$. For incomplete lists the rank function becomes
\begin{equation}
	\widetilde R_l(X_p) = \begin{cases}
	     \{R_l(X_p)\} & \text{for } X_p\in \Lambda_l\\
	     \{d_l+1,\dots,P\} & \text{for} X_p \not\in \Lambda_l
        \end{cases}
\end{equation}
where we only know that the rank for the unobserved items in list $l$
must be larger than the largest rank observed in that list.

The agreement, $A(X_p)$, cannot be computed directly for all
predictors in the presence of incomplete lists because the exact rank
for some items will be unknown. Also, recall that the rankings within
a single list are not independent since each rank must appear exactly
once in each list. Thus, we cannot simply assign the same number
(\eg, the mean of the unassigned ranks) to the missing items since
that would result in \emph{less} variation of the ranks and hence less
variation of the agreement, and it would artificially introduce a
(downward) bias of agreement for items that are missing in multiple
lists.

@@

\begin{algorithm}
\caption{Sequential rank agreement algorithm for incomplete lists}
\label{sra-algorithm}
\begin{algorithmic}[1]
  \Procedure{Incomplete list rank agreement}{} \State Let $B$ be the number
  of permutations to use \For{each $b \in B$} \For{each list
    $l \in L$}
  \State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1.7cm}{Permute
    the unassigned ranks, $\{d_l+1, \ldots, P\}$, and assign them
    randomly to the items \emph{not} found in the list, i.e.,
    $\Lambda^\complement=X\setminus\Lambda_l$. Combine the
    result with $\Lambda_l$ to fill out the list.}
\EndFor
\State \parbox[t]{\dimexpr\linewidth-\algorithmicindent-1cm}{Let sra($b$) be the sequential rank agreement computed from the
filled out lists.}
\EndFor
\State Return element-wise averages across all $B$ permutations of sra($b$).
\EndProcedure
\end{algorithmic}
\end{algorithm}

The proposed approach is based on two assumptions: 1) that the most
interesting items are found in the top of the lists, and 2) that the
ranks that are missing from the lists provide so little information that it is reasonable
to assume that they can be represented by a random order. The first
assumption is justifiable because we have already accepted that it is
reasonable to rank the items in the first place. The second assumption
is fair in the light of the first assumption provided that we have a
``sufficiently large'' part of the top of the lists available.

When the two assumptions are satisfied then it is clear that the
interesting part of the sequential rank agreement curves is restricted
to the depths where the number of items without ranks available is
low. Generally, without additional prior knowledge about the
underlying distribution of the ranks it seems reasonable to restrict
attention of the sequential rank agreement to lists no longer than the
longest of the $L$ lists examined.

Like for fully observed lists we generally expect the sequential rank
agreement to start low and then increase unless the lists are
completely unrelated (in which case the sequential rank agreement will
be constant at a high level) or if the lists mostly agree on the
ranking (in which case the sequential rank agreement will also be
constant but at a low level). For incomplete ranked lists we also
expect a changepoint around the depth where the lists are
become incomplete. This is an artefact stemming from the fact that we assume
that the remainder of the lists can be replaced by a simple
permutation of the missing items.

The two right-most plots of Figure~\ref{fig:example1} show the impact
of restricting the Golub data such that only top-20 lists are
available instead of full lists of length 1000 (20 was chosen to
resemble the list lengths that might be published in a
manuscript). The sequential rank agreement increases much quicker
because the incomplete lists introduce more noise in the estimation of
the agreement, but it is still possible to see that the top of the
list has a sequantial rank agreement that is not substantially
different from the full lists.

@@

The first reference hypothesis is
\begin{eqnarray}
H_0  & : &  \text{The list rankings correspond to completely randomly}\label{eq:permutationHypothesis}\\
       &  & \text{permuted lists}\nn
\end{eqnarray}
which not only assumes that there is no information in the data on
which the rankings are based but also that the methods used to provide
the rankings are completely independent.

Alternatively, we can remove the restriction on the independence among
the methods used to generate the individuals rankings in the $L$ lists
by only requiring that there is no information contained in the ranking but that
the rankings are all based on applying the method/approaches to the
same data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

This alternative null hypothesis addresses the fact that some ranking
methods are more likely to provide similar rankings of the same data
because the ranking methods focus on the same features of the data
rather than because of any information contained in the data.

@@

$H_0$ is a quite unrealistic null hypothesis but we can easily obtain
realizations from that null hypothesis simply by permuting the items
within each list and then computing the sequential rank agreement for
the permuted lists. In the fully observed case each experiment
contains $L$ lists of random permutations of the items in $X$. For the
incomplete case we first permute the items $X_1,\dots,X_P$ and then assign missing ranks
for list $l$ from  $d_l$ to $P$ (\ie, each list so it has the same number of
observed rankings as was observed for list $l$ in the original
dataset). The sequential rank agreement curve from the original
lists can then be compared to, say, the pointwise 95\% quantiles of
the observed rank agreements obtained under $H_0$.

@@

Figure~\ref{fig:example1} shows the empirical distributions of
sequential rank agreement under $H_0$ and $\widetilde H_0$ each based
on $400$ permutations of the Golub data described in Section
\ref{sec:amfol}. Not surprisingly, the sequential rank agreement under
$\widetilde H_0$ is lower than the sequential rank agreement under
$H_0$ because the four methods used to rank the data ($t$ test,
logistic regression, elastic net, and MIC) generally tend to identify
(and rank) similar predictors even if there are only spurious
associations. The two bottom panels in Figure~\ref{fig:example1} also
indicate that the observed sequential rank agreement (the black line)
is better than what would be expected by chance for data that contain
no information since it lies below the reference areas. The incomplete
data also suggests that there may be at most 1 or 2 ranked items
towards the top of the lists that yield a result better than what
would be expected (the bottom-right plot).

@@

Corollary \ref{thecor} indicates that we can use the threshold
function $\widehat{q}_L$ estimated under the null hypothesis as
discussed in the previous section as a limiting threshold function for
inferring the depth $d$, where the observed sequential rank agreement
first crosses the threshold of the null threshold, \ie, the depth
until which the observed ranked lists are in better agreement than
expected under the null hypothesis. In that sense the threshold
function serves the same role as the limits of agreement in method
comparison studies, except that the threshold function is not constant
but can accommodate the changing nature of the number of items used
for the computation of the sequential rank agreement for a given depth.

@@

The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce incomplete lists depending on
the value of the penalty parameter.

@@

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we
see that a large proportion of them are numerically very close to zero
and have almost equal absolute value. This is a general feature of the
Ridge models in this dataset and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient
values below a given threshold thereby introducing an artificial
incompletion of the lists.  For the Ridge models tuned with the
deviance criterion, Figure \ref{fig:app2} (left) shows the the
sequential rank agreement where for each of the 1000 trained models
the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated using Algorithm
\ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (right panel) is shown for
comparison.  Even though the number of predictors with missing ranks is very
small compared to the total number of predictors, the effect on the
sequential rank agreement is substantial and with the artificial
censoring the shape of the curves is as expected, starting low and
then increasing.

Looking at the agreement curves for the Lasso models in Figure
\ref{fig:app1} (right) we clearly see the effect of the sparsity
inducing penalization giving rise to incomplete lists. These curves
were similarly calculated using \ref{sra-algorithm} and $1000$ random
permutations. Under the deviance optimization criterion the median
number of non-zero coefficients was 33 (range 16 to 50) and for the
class accuracy criterion 14 (range 4 to 56). These values correspond
to the list depths where the agreement curves become constant as a
result of the subsequent censoring.

@@

\begin{figure}[htbp]
\begin{center}
\includegraphics[width=.49\textwidth]{pics/ridgeCensoring}%
\includegraphics[width=.49\textwidth]{pics/aucPlot}
\end{center}
\caption{Left panel: Sequential rank agreement for Ridge regression
  obtained by artificially censoring predictor ranks when their absolute
  coefficient values are lower than the 0.1\% quantile. Right panel:
  Box plots of AUC values across the $1000$ sub-samples with respect
  to the known class labels of the DACOVA data.}
 \label{fig:app2}
\end{figure}

@@

Lasso shows similar agreement for ranking the risk predictions than
PLS-DA (except for the AUC criterion), and poorer agreement for
ranking predictors. This reason for the latter is the high
auto-correlation between the intensities in the mass spectra which
leads to collinearity issues in the regression models. It is
well-known that variable selection with the Lasso penalty does not
perform very well when the predictors are highly correlated. The
collinearity does, however, not affect the agreement of the risk
predictions, as seen on the left panel of Figure \ref{fig:app1}, since
it is not so important which specific variable that gets selected from
a group of highly correlated predictors when the purpose is risk
predictions.

@@

We present results from a simulation study where we investigated the
small sample properties of the sequential rank agreement and compared
it to the topK method \citep{hall:schi:2012} with respect to list
depth agreement. For the sequential rank agreement we use the function
\texttt{sra} of the \texttt{R} package \texttt{SuperRanker} with the
\texttt{"mad"} option (median absolute distance). To define the depth
of agreement we set a constant threshold function to an integer $q$
and report the first crossing point, i.e., the smallest list depth
where sra exceeds $q$. For topK we use the function
\texttt{j0.multi} which is implemented in the \texttt{R} package
\texttt{TopKLists}. Specifically, we set the tuning parameter
\texttt{v} of \texttt{j0.multi} to the value 6 and the window
parameter \texttt{d} to $q$ and report the output parameter
\texttt{maxK} as the depth of agreement. Thus, to make this comparison
we assume that the first crossing point of sra and the result of the
topK method measure the same underlying feature.

@@

\section{Discussion}
In this article we address the problem of comparing ranked lists of
the same items. Our proposed method can handle both the situation
where the underlying data to generate the ranked lists are available
and the situation where the only available data is the actual ranked
lists. In addition, incomplete ranked lists where only the ranks of the
top $k$ ranked items are known can be accommodated as well. The
proposed agreement measure can be interpreted as the average distance
between an item's rank and the average rank assigned to that item
across lists. Thus the measure determines how well the lists agree on
the ranks of a specific item.

@@

Finally, we have --- whenever possible --- used all available ranks
from the lists.  We could choose to restrict attention to the rank of
items which show evidence for significance in their models. That would
ensure that there would be put less emphasis on the agreement of the
non-significant items and it would be easier to identify a change in
agreement among the items that were deemed to be relevant. In our
application section we have successfully introduced such an artificial
censoring for the predictor rankings obtained with ridge regression.
